---
title: "NYPD Shooting Incidents"
author: ''
date: "3/25/2022"
output:
  pdf_document: default
  html_document: default
---
```{r libraries,message = FALSE}
library(tidyverse)
library(lubridate)
library(hms)
library(ggplot2)
library(patchwork) #to display to plots together
library(readxl)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Import Data Set from URL
Read the NYPD Shooting Incident Data (Historic) from <https://catalog.data.gov/dataset> and store it in **nypd** 
List of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year.

This is a breakdown of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year. This data is manually extracted every quarter and reviewed by the Office of Management Analysis and Planning before being posted on the NYPD website. Each record represents a shooting incident in NYC and includes information about the event, the location and time of occurrence. In addition, information related to suspect and victim demographics is also included. This data can be used by the public to explore the nature of shooting/criminal activity.

```{r dataimport}
nypd <- read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv")
```
### Tidy and Transform your data

#### First we deal with missing or missnamed data. 

We will set all the missing or unknown data to Na or Unknown. As well as the data that makes no sense like perp age group 1020 or 940. 

We will start with the Location Description, as is the first one that has empty rows transforming the empty ones to 'NONE' that is one of the options. 

```{r location_desc_empty}
nypd[nypd$LOCATION_DESC == "",]$LOCATION_DESC <- "NONE"
```
For PERP_AGE_GROUP the available options are: 
["<18", "18-24", "25-44", "45-64", "65+", "UNKNOWN"] options that aren't an age group or "","UNKNOWN" I will consolidate change to "UNKONWN".
For PERP_SEX will change empty ones for "U" as in unknown same for PER_RACE empty ones will say "UNKNOWN
```{r PERP }
nypd[!(nypd$PERP_AGE_GROUP %in% c("<18", "18-24", "25-44", "45-64", "65+", "UNKNOWN")),]$PERP_AGE_GROUP <- "UNKNOWN"
nypd[nypd$PERP_SEX == "" ,]$PERP_SEX <- "U"
nypd[nypd$PERP_RACE == "" ,]$PERP_RACE <- "UNKNOWN"
```
The rest of the columns don't seem to have any missing values.

#### Second we tranform data type and get rid of columns we dont need

Most of the data is stored as character but is more useful as categorical as it comes from categories like boroughs, race, age group, and if it was flagged as murder (the victim died) we change it to logical (TRUE, FALSE). We also get rid of the Lon_Lat column as we already have the long and lat separately. The location and longitude and latitude could be needed in case we want to find location hot-spots, locate the shootings in a map or find the distance between shootings so I will keep such data. 

```{r clean}
nyp <- nypd %>%
    mutate(OCCUR_DATE = mdy(OCCUR_DATE), OCCUR_TIME = as_hms(OCCUR_TIME), BORO = as.factor(BORO), PRECINCT=as.factor(PRECINCT), JURISDICTION_CODE = as.factor(JURISDICTION_CODE), LOCATION_DESC = as.factor(LOCATION_DESC),MURDER_FLAG = as.logical(STATISTICAL_MURDER_FLAG), PERP_AGE_GROUP = as.factor(PERP_AGE_GROUP), PERP_SEX = as.factor(PERP_SEX), PERP_RACE= as.factor(PERP_RACE), VIC_AGE_GROUP = as.factor(VIC_AGE_GROUP), VIC_SEX = as.factor(VIC_SEX), VIC_RACE = as.factor(VIC_RACE)) %>% 
    select(-c(STATISTICAL_MURDER_FLAG,Lon_Lat))
```
### Summary of the data
Check for missing data.
```{r summary}
summary(nyp)
```
### Visualizations. 
First lets plot  the total number of shootings and murders by gun violence per year in NYC and how it changed over the last 15 years. Including how lethal this shootings were. 
```{r Yrl shootings,fig.height= 8}

nyp$Yr <- year(nyp$OCCUR_DATE)
yrl <- nyp %>%
  group_by(Yr) %>%
  summarize(shootings = n(), murders_gun = sum(MURDER_FLAG))
yrl$deadly_per = yrl$murders_gun/yrl$shootings

p1 = ggplot(yrl, aes(x=Yr,y = shootings)) + geom_point(color = 'blue' ) + geom_line(color = 'blue' ) + labs(x = NULL,y = "# of Shootings", title = "Gun violence in NYC from 2006 to 2020") + 
  scale_x_continuous(breaks = seq(2006,2021,2), labels = NULL)

p2 = ggplot(yrl, aes(x=Yr,y=murders_gun)) + geom_point(color = 'purple' ) +geom_line(color = 'purple' ) + labs (x = NULL, y = "# deads by gun violence") +scale_y_continuous(breaks = c(200,300, 400)) + 
  scale_x_continuous(breaks = seq(2006,2021,2),labels = NULL)

p3 = ggplot(yrl, aes(x=Yr,y = deadly_per)) + geom_point(color = 'red' ) + geom_line(color = 'red' ) + labs(x = "Year",y = "% of Shootings that are lethal") + scale_y_continuous(breaks = seq(0.15,0.25,0.01)) +
  scale_x_continuous(breaks = seq(2006,2021,2))

layout <- c(area(1,1, 3, 1),area(4,1,5, 1), area(6,1,7,1))
p1 + p2  + p3 +plot_layout(design = layout)
```

We can see that gun violence dropped slowly from 2006 to 2011 and then had a big drop until 2017, was stable until 2019 and then skyrocket back to 2011 levels. In 2020  the number of shootings doubled the ones of 2019.
But at least the lethality did not increase. 

I wanna see when this trend started so will analyze the year 2020 by month and compare it to previous years. 

```{r 2019-2020 comparison, message = FALSE , warning = FALSE}
ny1920 <- nyp %>%
  filter(Yr > 2015) %>%
  mutate("Month"= month(OCCUR_DATE), "Year" = as.factor(Yr)) %>%
  group_by(Year,Month) %>%
  summarize(shootings = n(), murders_gun = sum(MURDER_FLAG))


f1920 <- ggplot(ny1920) + geom_point(aes(x=Month, y=shootings, colour=Year)) 
f1920 + geom_line(aes(x=Month, y=shootings, colour=Year)) + scale_x_continuous(breaks = seq(1,12), labels =c('JAN',"FEB", "MAR","APR","MAY","JUN","JUL","AGO","SEP","OCT","NOV","DIC")) + labs(x = "Month",y = "# of Shootings", title = "Number of Shootings in NYC per Month")
```

We can see a yearly trend where shootings increase during the warmer months of the year. However the shootings of 2020 are almost 3 times more than on previous summers. Besides the 2020 peak that started in May, there is also a peak in August 2016. 

Let's check shootings by year by borough, to see if such increase was in all NYC or concentrated in some boroughs. First we need a column with the year and then we can group them and find how many shootings happened per year.  
```{r shootings_per_year_Boro, message = FALSE , warning = FALSE}

nyp_by_b <- nyp %>% 
  group_by(BORO,Yr) %>% 
  summarize(shootings = n(), murders = sum(MURDER_FLAG))
shb <- ggplot(nyp_by_b) + geom_point(aes(x=Yr, y=shootings, colour=BORO)) 
shb + geom_line(aes(x=Yr, y=shootings, colour=BORO)) + scale_x_continuous(breaks = seq(2006,2021,2)) +
  labs(x = "Year",y = "# of Shootings", title = "Number of yearly Shootings per Borough")
```


The shootings in every borough increased but it seems like the stepper increase was in Brooklyn that already had the highest gun violence rates in the city.

This raised the question of if the shootings per population are also the highest in Brooklyn or because Brooklyn is the borough with the highest population in the city it has the highest number of shootings. 

So to give a fair comparison between boroughs I will find the number of shootings per 10,000 habitants. However let's keep in mind that a perpetrator that lives out of a specific borough can shot a gun in the borough.

Because the increase was between 2019 and 2020, we will look at how much the murder rate per borough and borough population increased. 
 
The data of The population by borough comes from <https://data.cityofnewyork.us> and I will use the data for 2020. 

```{r population data,message = FALSE , warning = FALSE }

#Get NYC population by borough
pop <- read_csv("https://data.cityofnewyork.us/api/views/xywu-7bv9/rows.csv") 

#Keep only 2020
population <- pop %>% 
  select(Borough,`2020`) %>%
  rename('population'='2020', 'BORO'='Borough') %>%
  filter(BORO!= "NYC Total")%>%
  mutate(BORO = toupper(BORO))

#From the shootings data keep only 2019-2020 group it by Year and borough and rearange data to have a row for each borrough
ny1920b <- nyp %>%
  filter(Yr > 2018) %>%
  mutate("Year" = as.factor(Yr)) %>%
  group_by(Year, BORO) %>%
  summarize(shootings = n()) %>%
  pivot_wider(names_from = Year, values_from = shootings,names_prefix = 'Y')  

#Add population data to shootings data frame
ny1920b <-left_join(ny1920b,population,by = 'BORO')
#Get the number of shootings per 10,000 habitants and reorder data frame
nyper <- ny1920b %>%
  mutate('S_2019_per_10000'=10000*Y2019/population,'S_2020_per_10000'=10000*Y2020/population ) %>%
  rowwise()%>%
  mutate(mymean = mean(c(S_2019_per_10000,S_2020_per_10000))) %>%
  arrange(mymean) %>%
  mutate(Borough = factor(BORO, BORO), Increase =S_2020_per_10000/S_2019_per_10000 )

#Plot data using lolipop
ggplot(nyper)+
  geom_segment(aes(x = Borough, xend = Borough, y =S_2019_per_10000, yend=S_2020_per_10000), color = 'grey' )+
  geom_point(aes(x = Borough,  y =S_2019_per_10000, color = factor(2019)))+
  geom_point(aes(x = Borough,  y =S_2020_per_10000, color = factor(2020)))+
  ylim(0,4)+
  scale_y_continuous(breaks= seq(0,4,0.5), limits=c(0,4))+
  labs(x= "",y="Shootings per 10,000 habitants",title="Increase in shootings per 10,000 habitants between 2019 and 2020",colour = "Year")+ scale_fill_manual("Year",values=c("2019","2020"))
```


We can see that even Brooklyn is the Borough with the most total number of shootings, the Bronx has more shootings per 10,000 people both in 2019 and 2020.

The model for this data would be that for each Borough the number of shooting of 2020 were 2 times the number of shootings of 2019 in that same borough.  


### Conclusions and bias Identification

The number of shootings registered in NYC had an steep increase in 2020 and reached the levels of 10 years ago and double than the values from 2019. 
In the last five years there are more shootings during the summer than in any other season.
Of the 5 Boroughs Brooklyn has the highest number of shootings per year, but the Bronx has the highest per capita. Staten Island has the least amount of total shootings and shootings per capita. 

I avoided trying to explain  why the number of shootings increased either in 2020 or in August of 2016 as I would be probably biased. I also didn't show any information of the gender or race or the perps as I saw in the summary that a lot of it said unknown and from the ones it did have information I don't how such information was collected and if it was biased when collected.  My bias might be that I did not look into this information to try to avoid being biased. 